\chapter*{Acknowledgments}

To my beloved Mum, Linda Wood, who passed away while I was writing this thesis. Words cannot express my gratitude for the unwavering emotional support you offered me over the years. You were my biggest fan and my harshest critic, and you will be forever missed.

Thank you to my family: Dave, Jo, and Freya. I am grateful for your constant support and for patiently enduring my endless discussions about Artificial Intelligence (AI) at the dinner table. Your encouragement throughout my PhD journey has meant the world to me.

To my partner of six years, Millie, thank you for your love, invaluable advice, and enduring support, especially through the lean years of student life.

Beyond my immediate circle, I want to thank my friend Chris for being a constant sounding board for my rants about AI and for his encouragement that this work could have a positive impact. I also extend my gratitude to the wider group of friends and colleagues who provided camaraderie along the way: Andrew (x2), Anton, Eduardo, Mashfaq, Christian, Hayden, Jordan, Rimas, Carl, Kaan, Junhao, Peng, and Binh.

My sincere thanks go to Alastair Ross from AgResearch for generously providing the REIMS dataset that is the core subject of this thesis, as well as for his valuable feedback and essential insights into the practical applications of this work.

Finally, I wish to extend my deepest gratitude to my supervisors, without whose guidance this thesis would not have been possible: Dr. Bach Nguyen, Prof. Bing Xue, Prof. Mengjie Zhang, and Dr. Daniel Killeen. I am indebted to Bach for his meticulous attention to detail and deep expertise. I thank Bing for her consistent emotional support and inspiring dedication to her field. I am grateful to Meng for his impeccable grammar and breadth of knowledge, and to Daniel for his invaluable feedback and crucial insights into the chemical world.

This thesis was financially supported by the University Maori PhD Scholarship and the Cyber-Marine MBIE Endeavor Research Program under contract C11X2001 for both tuition fees and stipends.